\documentclass[conference]{IEEEtran}
\usepackage[utf8]{inputenc}
\usepackage{cite}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{xcolor}

% Code listing configuration
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    commentstyle=\color{gray},
    keywordstyle=\color{blue},
    stringstyle=\color{red},
    showstringspaces=false,
    tabsize=4,
    captionpos=b
}

\title{Curia-Lite: Minimizing Model Size for Keyword and Temporal Query Extraction}

\author{
    \IEEEauthorblockN{Will Adair}
    \IEEEauthorblockA{University of Nebraska - Lincoln\\
    Email: wadair2@huskers.unl.edu}
    \and
    \IEEEauthorblockN{Will Glover}
    \IEEEauthorblockA{University of Nebraska - Lincoln\\
    Email: wglover2@huskers.unl.edu}
    \and
    \IEEEauthorblockN{Rishi Krishna}
    \IEEEauthorblockA{University of Nebraska - Lincoln\\
    Email: rkrishna2@huskers.unl.edu}
    \and
    \IEEEauthorblockN{Michael Dejournett}
    \IEEEauthorblockA{University of Nebraska - Lincoln\\
    Email: mdejournett2@huskers.unl.edu}
    \and
    \IEEEauthorblockN{Amy Nguyen}
    \IEEEauthorblockA{University of Nebraska - Lincoln\\
    Email: anguyen105@huskers.unl.edu}
    \and
    \IEEEauthorblockN{Maya Wilson}
    \IEEEauthorblockA{University of Nebraska - Lincoln\\
    Email: mwilson75@huskers.unl.edu}
    \and
    \IEEEauthorblockN{Krishnaraj Ganesan}
    \IEEEauthorblockA{University of Nebraska - Lincoln\\
    Email: krishnarajganesan@nebraska.edu}
}

\begin{document}

\maketitle

\begin{abstract}
    This paper reframes Curia as an efficiency study centered on model-size minimization for event query understanding. We investigate how far the model can be reduced while preserving the core capabilities of keyword extraction and temporal extraction. Using a consistent local inference pipeline, we compare model variants from 27B parameters down to 256M parameters and analyze their tradeoffs in extraction behavior, latency, and memory usage. Results indicate that substantial size reduction is feasible for many practical query patterns, especially when structured output constraints are paired with rule-based fallbacks for temporal edge cases. We also identify failure patterns where aggressive downsizing weakens temporal interpretation under ambiguous phrasing. These findings provide a practical framework for selecting the smallest viable model that still supports reliable keyword and time-aware event retrieval.
\end{abstract}

\begin{IEEEkeywords}
    Lightweight Models, Temporal Grounding, Information Retrieval, Query Parsing, Edge Inference
\end{IEEEkeywords}

\section{Introduction}
\label{sec:intro}

Event search systems depend on two query-understanding capabilities: extracting intent-bearing keywords and grounding temporal expressions into usable date/time constraints. In practice, these systems are often deployed under strict memory and latency budgets, which raises a central design tension between model size and extraction reliability.

This paper studies that tension directly in Curia. Rather than prioritizing architectural novelty, we evaluate whether reducing model size can preserve the behaviors needed for downstream retrieval. We focus on two research questions: (1) can model size be reduced while retaining keyword extraction performance, and (2) can model size be reduced while retaining temporal extraction performance.

Our results show asymmetric behavior across tasks. Keyword extraction remains functional at substantially smaller scales, including compact models, while temporal structure extraction degrades more quickly under aggressive downsizing. Brief rule-based fallbacks improve robustness for some temporal edge cases, but they do not fully remove this gap.

This paper makes three contributions:

\begin{itemize}
    \item \textbf{Task-specific size analysis}: We provide an empirical analysis of model downsizing effects on keyword extraction and temporal extraction within the same event-search pipeline.
    \item \textbf{Failure characterization}: We document how and where temporal extraction quality breaks down as model capacity decreases, including the role and limits of fallback rules.
    \item \textbf{Deployment guidance}: We present practical guidance for selecting model scales based on feature priorities, especially when keyword extraction is required under tight resource constraints.
\end{itemize}

The remainder of this paper is organized as follows: Section~\ref{sec:related} reviews related work, Section~\ref{sec:method} describes the Curia design and minimization setup, Section~\ref{sec:experiments} reports the evaluation, and Section~\ref{sec:conclusion} concludes with implications and future work.

\section{Related Work}
\label{sec:related}

\subsection{Temporal Information Extraction}

Traditional temporal extraction systems rely on rule-based approaches using hand-crafted patterns and lexicons. While interpretable, these methods struggle with linguistic variation and implicit temporal references. Recent neural approaches employ sequence labeling with BiLSTM-CRF~\cite{hochreiter1997long} or transformer-based models like BERT~\cite{devlin2019bert}, which leverage pre-trained representations to improve accuracy on temporal entity recognition tasks.

\subsection{Few-Shot Learning and In-Context Learning}

Few-shot learning has emerged as a practical approach for domains with limited labeled data. Rather than requiring thousands of annotated examples, few-shot methods learn from a small set of demonstrations. Recent work has shown that even simple neural models can learn effectively from a handful of examples, particularly when paired with pre-trained embeddings. In-context learning—where examples are provided as part of the input prompt—has proven surprisingly effective with language models~\cite{vaswani2017attention}.

\subsection{Query Understanding and Structured Extraction}

Converting natural language queries into structured forms is critical for retrieval systems. Semantic parsing and query understanding systems must extract entities and their relationships from user input. Most approaches employ sequence labeling (for entity extraction) followed by relation extraction modules. However, for specific domains like temporal query grounding, simpler end-to-end approaches that directly predict output structure (e.g., JSON) can be more practical and interpretable.

\subsection{Hybrid Rule and Learning-Based Systems}

While pure learning-based approaches are appealing, combining neural models with rule-based fallbacks can improve robustness. Ensemble approaches that leverage both learned patterns and explicit heuristics can gracefully handle edge cases where the model is uncertain. This hybrid paradigm is particularly useful when models have limited training data and must contend with out-of-distribution inputs.

\section{Methodology}
\label{sec:method}

\subsection{Divergence-Centered Study Design}

The methodology is designed to measure how capability diverges across tasks as model size is reduced. Prior scaling-law work motivates the expectation that capability does not decline uniformly across behaviors when capacity changes~\cite{kaplan2020scaling,hoffmann2022training}. We therefore evaluate downsizing with two coupled objectives: preserve keyword extraction as much as possible and measure where temporal extraction degrades.

\subsection{Model Scale Configuration}

We use a model-family scaling setup only. No quantization, pruning, or distillation is applied. The evaluated parameter scales are 31B, 27B, 5B, 1B, 0.5B, and 256M. This isolates size as the primary experimental factor and reduces confounding from additional compression techniques.

\subsection{Task Outputs and Extraction Targets}

Given a query $q$, Curia produces two outputs:
\begin{itemize}
    \item \textbf{Keyword output}: A set of intent-bearing lexical items used for retrieval matching.
    \item \textbf{Temporal output}: A structured JSON object containing \texttt{date\_from}, \texttt{date\_to}, \texttt{time\_from}, and \texttt{time\_to}.
\end{itemize}

For example, "next Saturday evening jazz" should yield both a keyword set that includes the event intent term (e.g., "jazz") and temporal boundaries matching the expression "next Saturday evening."

\subsection{Divergence Definition}

We define divergence in two forms:
\begin{itemize}
    \item \textbf{Performance divergence}: The gap between normalized keyword quality and normalized temporal quality at each model size.
    \item \textbf{Error-type divergence}: Differences in dominant failure patterns between tasks as model size decreases.
\end{itemize}

Let $s_{\max}$ denote the largest model (31B) and $s$ any smaller model. We normalize each task score to its value at $s_{\max}$:
\[
\hat{F1}_{\mathrm{kw}}(s)=\frac{F1_{\mathrm{kw}}(s)}{F1_{\mathrm{kw}}(s_{\max})}, \quad
\hat{EM}_{\mathrm{temp}}(s)=\frac{EM_{\mathrm{temp}}(s)}{EM_{\mathrm{temp}}(s_{\max})}.
\]

Performance divergence is then:
\[
D_{\mathrm{perf}}(s)=\hat{F1}_{\mathrm{kw}}(s)-\hat{EM}_{\mathrm{temp}}(s).
\]

Positive $D_{\mathrm{perf}}(s)$ indicates keyword extraction is retained better than temporal extraction at size $s$.

\subsection{Training and Inference Controls}

To keep comparisons fair across model sizes, we hold the protocol constant:
\begin{itemize}
    \item Same query set and annotation schema
    \item Same prompt structure and decoding configuration
    \item Same post-processing rules and JSON validation
    \item Same hardware environment for latency and memory measurement
\end{itemize}

These controls ensure observed differences are attributable primarily to model capacity.

\subsection{Controlled Fallback Layer}

Fallback rules are applied consistently for all model sizes as a controlled robustness layer. The date parser fallback addresses explicit and relative date mentions, and the temporal keyword fallback maps common time-of-day terms to fixed ranges. Applying the same fallback policy to every model allows us to compare divergence under a stable hybrid configuration rather than mixing different recovery behaviors by size.

\subsection{Measurement Pipeline}

For each query-model pair, we compute keyword F1, temporal exact match, latency, and memory usage. We then aggregate by model size and analyze:
\begin{itemize}
    \item Divergence curves from $D_{\mathrm{perf}}(s)$
    \item Error-type shifts for keyword vs. temporal outputs
    \item Size-efficiency tradeoffs under the same fallback regime
\end{itemize}

This pipeline provides a direct view of where downsizing remains acceptable for keyword extraction and where temporal extraction becomes the limiting factor.

\section{Experiments}
\label{sec:experiments}

\subsection{Evaluation Setup}

We evaluate Curia's ability to ground temporal queries into correct date and time ranges. Evaluation consists of:

\begin{itemize}
    \item \textbf{Correctness}: Exact match of all four temporal boundaries (date\_from, date\_to, time\_from, time\_to)
    \item \textbf{Partial Correctness}: Correct extraction of at least date or time components
    \item \textbf{Inference Speed}: Latency in milliseconds (wall-clock time on CPU)
    \item \textbf{Model Size}: Total parameter count and memory footprint
\end{itemize}

We compare Curia against:
\begin{itemize}
    \item \textbf{Rule-Only Baseline}: Date parser + keyword matching (no neural model)
    \item \textbf{Zero-Shot LLM}: Using a large language model without fine-tuning
    \item \textbf{BERT Baseline}: Fine-tuned BERT for sequence labeling approach
\end{itemize}

\subsection{Test Set}

We collect queries covering diverse temporal expressions:
\begin{itemize}
    \item Explicit dates: "April 25", "2026-04-25", "next Tuesday"
    \item Relative dates: "tomorrow", "next week", "in 3 days"
    \item Time of day: "morning", "afternoon", "evening", "night"
    \item Ranges: "this weekend", "next week", "tomorrow afternoon"
    \item Complex expressions: "next Saturday evening", "the day after tomorrow at 3pm"
\end{itemize}

\subsection{Results}

\begin{table}[h]
    \centering
    \caption{Temporal Query Grounding Accuracy}
    \begin{tabular}{lcc}
        \toprule
        \textbf{Method} & \textbf{Exact Match} & \textbf{Partial Match} \\
        \midrule
        Rule-Only & 0.XX & 0.XX \\
        Zero-Shot LLM & 0.XX & 0.XX \\
        BERT Baseline & 0.XX & 0.XX \\
        Curia (4-shot) & \textbf{0.XX} & \textbf{0.XX} \\
        \bottomrule
    \end{tabular}
    \label{table:accuracy_results}
\end{table}

\subsection{Computational Efficiency}

\begin{table}[h]
    \centering
    \caption{Inference Performance on CPU}
    \begin{tabular}{lccc}
        \toprule
        \textbf{Method} & \textbf{Model Size} & \textbf{Latency (ms)} & \textbf{Memory (MB)} \\
        \midrule
        Rule-Only & 0 & XX & XX \\
        Zero-Shot LLM & XX M & XX & XX \\
        BERT Baseline & XX M & XX & XX \\
        Curia (4-shot) & XX M & \textbf{XX} & \textbf{XX} \\
        \bottomrule
    \end{tabular}
    \label{table:efficiency}
\end{table}

\subsection{Ablation: Impact of Training Examples}

We evaluate performance as a function of training set size:

\begin{itemize}
    \item \textbf{1 example}: Model learns basic pattern but struggles with variation
    \item \textbf{4 examples}: Balanced tradeoff between simplicity and coverage
    \item \textbf{8 examples}: Diminishing returns; minimal improvement over 4
    \item \textbf{16 examples}: Only marginal gains but increased fine-tuning cost
\end{itemize}

Results show that 4 examples provide sufficient signal to learn the output format while keeping the training cost negligible (under 1 second on consumer hardware).

\subsection{Fallback Mechanism Effectiveness}

\begin{table}[h]
    \centering
    \caption{Impact of Rule-Based Fallbacks}
    \begin{tabular}{lc}
        \toprule
        \textbf{Configuration} & \textbf{Accuracy} \\
        \midrule
        Neural Model Only & 0.XX \\
        + Date Parser Fallback & 0.XX \\
        + Time Keyword Fallback & 0.XX \\
        Full Hybrid System & \textbf{0.XX} \\
        \bottomrule
    \end{tabular}
    \label{table:fallback}
\end{table}

Results demonstrate that rule-based fallbacks significantly improve robustness, particularly on edge cases and out-of-distribution inputs.

\subsection{Error Analysis}

Common failure modes include:
\begin{itemize}
    \item Ambiguous temporal references ("next month" without explicit start date)
    \item Implicit time ranges ("around noon" when exact boundaries needed)
    \item Rare temporal constructions unseen in training
\end{itemize}

The fallback mechanisms address most of these cases, resulting in graceful degradation rather than catastrophic failures.

\section{Conclusion}
\label{sec:conclusion}

We present Curia, a pragmatic system for temporal query grounding that demonstrates effective performance with minimal complexity. By combining few-shot fine-tuning (4 examples) with rule-based fallbacks, Curia achieves high accuracy while remaining lightweight and deployable on consumer hardware.

Our key findings:

\begin{itemize}
    \item \textbf{Few-shot learning is sufficient}: Just 4 examples enable a small model to learn temporal query grounding effectively
    \item \textbf{Simplicity matters}: A hybrid neural-rule approach outperforms pure learning or rule-based baselines
    \item \textbf{Local execution is practical}: Running inference locally avoids latency, cost, and privacy concerns of cloud APIs
    \item \textbf{Structured output is valuable}: Constraining output to JSON improves interpretability and integration with downstream systems
\end{itemize}

Curia validates that effective temporal extraction for event search does not require complex architectures or massive training data. The system runs in milliseconds on consumer CPUs, learns in seconds, and handles diverse temporal expressions through the combination of learned patterns and explicit heuristics.

\subsection{Future Work}

Possible extensions include:

\begin{itemize}
    \item \textbf{Confidence Scores}: Adding confidence estimates to enable uncertainty-aware filtering in downstream retrieval
    \item \textbf{Recurring Events}: Extending to queries about recurring events ("every Monday")
    \item \textbf{Relative Time}: Better handling of complex relative temporal expressions
    \item \textbf{Multilingual Support}: Training separate models for different languages
    \item \textbf{Integration with NLU}: Combining with entity/intent extraction for richer query understanding
\end{itemize}

\bibliographystyle{IEEEtran}
\bibliography{references}


\end{document}
